<!-- question-type: inclass -->
### Exercise 5: Joins with real data

You now know what `left_join()`, `inner_join()` and `anti_join()` do, and why a join key has to match the observational unit. In this exercise you will apply all of that to the full ASX dataset.

Run the code below to load and prepare the P/E ratio data.

```{r}
#| eval: false

pe_data <-
    read_csv("data/pe.csv") |>
    select(gvkey, fyear, pe) |>
    filter(!is.na(pe)) |>
    arrange(gvkey, fyear)
```

**(a)** Before joining anything, work out what you are joining.

* What does one row of `asx_prices_tidy` represent? How many variables together identify a row/observation?
* What about for `pe_data`?
* Which variables should you therefore join on?

**(b)** Complete the code below to add the `pe` variable from `pe_data` to `asx_prices_tidy`.

```{r}
#| eval: false
asx_with_pe <-
    asx_prices_tidy |>
    left_join(
        pe_data,
        by = YOUR_CODE_HERE
    )
```

Did the join add rows, columns, or both?

**(c)** Run the two other joins on the same pair of tables.

```{r}
#| eval: false
asx_with_pe_inner <-
    asx_prices_tidy |>
    inner_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )

missing_pe <-
    asx_prices_tidy |>
    anti_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

A join is not finished until you have checked what it did. Now compare the row counts.

```{r}
#| eval: false
nrow(asx_prices_tidy)
nrow(asx_with_pe)
nrow(asx_with_pe_inner)
nrow(missing_pe)
```

* Which two of these four numbers are equal, and why must they be?
* Which two add up to the total, and why must they?

**(d)** Using those row counts, what share of firm-year observations actually have a P/E ratio?

Is that higher or lower than you expected before you looked?

**(e)** A quarter of the data is a lot to lose without knowing why. Investigate `missing_pe` to work out where the gaps are.

Two things worth checking: whether the missing observations are concentrated in particular years...

```{r}
#| eval: false
missing_pe |>
    YOUR_CODE(fyear)
```

and whether they differ from the matched observations in some other way, for example, whether they differ in their earnings per share.
 
```{r}
#| eval: false
missing_pe |>
    summarise(
        n = n(),
        min_eps = YOUR_CODE_HERE(eps, na.rm = TRUE),
        max_eps = YOUR_CODE_HERE(eps, na.rm = TRUE),
        avg_eps = YOUR_CODE_HERE(eps, na.rm = TRUE)
    )
```

```{r}
#| eval: false
asx_with_pe_inner |>
    summarise(
        n = n(),
        min_eps = YOUR_CODE_HERE(eps, na.rm = TRUE),
        max_eps = YOUR_CODE_HERE(eps, na.rm = TRUE),
        avg_eps = YOUR_CODE_HERE(eps, na.rm = TRUE)
    )
```

Write one or two short findings, then suggest an explanation for what you found.

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Choose a join key by reasoning about the observational unit rather than by pattern-matching on column names.

* Apply `left_join()`, `inner_join()` and `anti_join()` to a realistic dataset, having met all three on the toy tables in Exercise 4.

* Use row counts as an arithmetic check on a join.

* Discover that missing data has structure, and form a hypothesis about its cause.

✅ **Core Concepts to Highlight**

The join key follows from the observational unit. Both tables are firm-year, so the key is `gvkey` and `fyear` together. This is the same argument students just made on three rows in Exercise 4(f).

The row counts are not decoration, they are an identity: `left` equals the original, and `inner` plus `anti` equals the original. If those do not hold, something is wrong.

P/E coverage here is only about 25%. This is the number that should make students uncomfortable, and part (e) is where that discomfort becomes a finding.

The missingness is not random. Mean EPS among the unmatched firm-years is negative, while among the matched it is positive. P/E is not meaningful when earnings are negative, so the data provider has almost certainly filtered those out.

This is the payoff of the whole tutorial: the joins are not just plumbing, they are how you find out whether your data can answer your question.

💬 **Suggested In-Class Prompts**

*"You have 9,484 rows and 2,407 P/E ratios. Do you still want to run this analysis?"*

*"Is the missing quarter missing at random? How would you check?"*

*"What is different about a firm with no P/E ratio?"*

*"If you reported the inner-join sample only, which firms would silently disappear from your report?"*

📌 **Common Misunderstandings**

Students may reach for `gvkey` alone out of habit. Explain the importance of joining with all of the keys that together identify one observation.

Students may report the coverage number without reacting to it. Push them to say whether 25% is acceptable.

Students may treat the missing values as a coding mistake on their part. They are a property of the supplied file.

Students may find the year pattern (fairly flat) and stop there. The year counts are the less interesting finding; the EPS comparison is the one that explains the mechanism.

Students may not think to compare the missing group against the matched group. Suggest computing the same statistic on both.

⏱️ **Timing**

**25 minutes.** Parts (a) to (d) are quick if Exercise 4 landed, so budget most of the time for (e) and discussion of selection effects and the importance of understanding your data.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

One row of `asx_prices_tidy` represents one firm in one year, so it takes `gvkey` and `fyear` together to identify it.

One row of `pe_data` also represents one firm in one year, giving the P/E ratio for that firm-year.

Because both tables are at the firm-year level, the join key must be `gvkey` and `fyear` together — exactly the situation from Exercise 4(f).

**(b).**

```{r}
asx_with_pe <-
    asx_prices_tidy |>
    left_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

The join added a column (`pe`) and left the number of rows unchanged, because each firm-year in `asx_prices_tidy` matched at most one firm-year in `pe_data`.

**(c).**

```{r}
asx_with_pe_inner <-
    asx_prices_tidy |>
    inner_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )

missing_pe <-
    asx_prices_tidy |>
    anti_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

```{r}
nrow(asx_prices_tidy)
nrow(asx_with_pe)
nrow(asx_with_pe_inner)
nrow(missing_pe)
```

`asx_prices_tidy` and `asx_with_pe` have the same number of rows (9,484), because a `left_join()` keeps every row on the left whether it matched or not.

`asx_with_pe_inner` (2,407) and `missing_pe` (7,077) sum to 9,484. They must, because every firm-year either found a match or did not, and the inner join keeps the first group while the anti join keeps the second.

**(d).**

```{r}
nrow(asx_with_pe_inner) / nrow(asx_prices_tidy)
```

Only about **25%** of firm-year observations have a P/E ratio. Roughly three quarters of the dataset has no `pe` value at all.

Most students expect a much higher number, and that gap is the point. A join that "worked" has still left us with a variable we can use for only a quarter of the sample.

**(e).**

First, check whether the gaps cluster by year.

```{r}
missing_pe |>
    count(fyear)
```

The missing observations are spread fairly evenly across years, from 993 in 2024 to 1,286 in 2022. So the problem is not that the P/E file covers a shorter time period.

That suggests looking at the firms rather than the years. Comparing EPS between the matched and unmatched groups is revealing.

```{r}
missing_pe |>
    summarise(
        n = n(),
        min_eps = min(eps, na.rm = TRUE),
        max_eps = max(eps, na.rm = TRUE),
        avg_eps = mean(eps, na.rm = TRUE)
    )
```

```{r}
asx_with_pe_inner |>
    summarise(
        n = n(),
        min_eps = min(eps, na.rm = TRUE),
        max_eps = max(eps, na.rm = TRUE),
        avg_eps = mean(eps, na.rm = TRUE)
    )
```

Two findings:

* Missing P/E ratios are spread across all six years, so this is not a time-coverage problem.
* Average EPS is **negative** (about −0.42) among the firm-years with no P/E ratio, but **positive** (about +0.34) among those with one. Every observation in the missing group has EPS of at most 2.05, while the matched group extends much higher.

The likely explanation is that P/E is not a meaningful statistic when earnings are negative — dividing a positive price by a negative EPS gives a negative ratio that cannot be interpreted as "years of earnings". The provider has almost certainly dropped or suppressed those firm-years, which also fits the many small, loss-making firms on the ASX.

This matters for analysis: the firms we can compute P/E for are systematically the profitable ones. Any conclusion drawn from the inner-join sample is a conclusion about profitable firms, not about the ASX as a whole.

:::

:::

:::

<!-- END PROFILE:r-solutions -->
