<!-- question-type: inclass -->
### Exercise 3: Tidying Data to Make Analysis Easier

**(a)** Look at the column names in `asx_prices_messy`, and answer the following:

* What two pieces of information are stored in these column names?
* What character separates those two pieces of information?

**(b)** We now want to reshape the stock price data so that each row corresponds to one firm in one year. Complete the starter code below to turn the data into a tidy format.

```{r}
#| eval: false
asx_prices_tidy <-
    asx_prices_messy |>
    pivot_YOUR_CODE(
        cols = YOUR_CODE_HERE,
        names_to = c(".value", "fyear"),
        names_sep = YOUR_CODE_HERE,
        values_drop_na = TRUE
    ) |>
    mutate(fyear = as.numeric(fyear))

asx_prices_tidy |>
    head(10)
```

**(c)** In Exercise 1, calculating the average share price across 2023 and 2024 took several steps. Use the tidy dataset to calculate the average share price across those two years.

```{r}
#| eval: false
asx_prices_tidy |>
    YOUR_CODE(fyear %in% c(YOUR_YEARS)) |>
    YOUR_CODE(avg_price = mean(YOUR_VARIABLE, na.rm = TRUE))
```

**(d)** Why is the calculation in part (c) easier to write using the tidy dataset than it was in Exercise 1?

What would we need to do if we added many more years of data?

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Identify the structure hidden inside compound column names and use that information to choose an appropriate reshaping method.

* Reshape wide stock price data into tidy long form using `pivot_longer()`.

* Use tidy data to carry out a simple summary far more efficiently than the wide version allowed.

* Explain why tidy data scales better when more years of data are added.

✅ **Core Concepts to Highlight**

Wide versus long data: when values that should be observations are spread across many columns, analysis becomes repetitive and harder to scale.

Column names can themselves contain data: here, names such as `price_2023` contain both a variable name and a year.

`pivot_longer()` can split those names into meaningful parts using `names_to` and `names_sep`.

Tidy data makes common tasks such as filtering and summarising much easier because the same variable is stored in one column rather than scattered across many.

💬 **Suggested In-Class Prompts**

*“What information is hidden inside a name like `eps_2024`?”*

*“Why is `pivot_longer()` the right choice here rather than `pivot_wider()`?”*

*“What would break first if this dataset expanded from two years to ten?”*

*“What is easier to change in a tidy workflow: the data or the code?”*

📌 **Common Misunderstandings**

Students may recognise the underscore but not yet see why that matters for reshaping.

Students may think tidying is mainly cosmetic rather than something that changes what analyses are easy to do.

Some students may try `pivot_wider()` because they associate ‘tidy’ with ‘more spread out’ or visually cleaner.

Students may not immediately see why one `filter()` on `fyear` is more scalable than manually referencing many separate year-specific columns.

⏱️ **Timing**

**15 minutes.** Part (b) gates everything later — if a group is still stuck on the pivot after 10 minutes, give them the answer rather than let them fall behind.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

The column names contain two pieces of information:

* the variable being measured, such as `price` or `eps`
* the financial year, such as `2023` or `2024`

These two pieces are separated by an underscore (`_`).

**(b).**

```{r}
asx_prices_tidy <-
    asx_prices_messy |>
    pivot_longer(
        cols = starts_with("price") | starts_with("eps"),
        names_to = c(".value", "fyear"),
        names_sep = "_",
        values_drop_na = TRUE
    ) |>
    mutate(fyear = as.numeric(fyear))

asx_prices_tidy |>
    head(10)
```

**(c).**

```{r}
asx_prices_tidy |>
    filter(fyear %in% c(2023, 2024)) |>
    group_by(fyear) |>
    summarise(avg_price = mean(price, na.rm = TRUE))
```

**(d).**

This version is easier because all share prices are stored in one `price` column and all years are stored in one `fyear` column. That means we can filter the years we want, then calculate the mean once, rather than referring to separate columns such as `price_2023` and `price_2024` individually.

If we added many more years of data, this tidy version would still work with the same basic code. We would just change the years inside `filter()`. In the wide version, we would need to keep adding more year-specific columns to the calculation, which becomes repetitive and harder to manage.

:::

:::

:::

<!-- END PROFILE:r-solutions -->
