<!-- question-type: inclass -->
### Exercise 4: Joins in Miniature

In Exercise 2 you used a `left_join()` to attach subindustry names to a list of firms. A `left_join()` is only one of several ways to combine two tables, and the choice between them matters.

Before we work with joins on the full ASX dataset, we will practise on two tiny tables where you can count every row by hand.

Run the code below to create them.

```{r}
firms_toy <- tribble(
    ~gvkey,  ~conm,           ~gsubind,
    "001",   "Alpha Mining",  15104020,
    "002",   "Bright Retail", 30101030,
    "003",   "Coral Energy",  10102020,
    "004",   "Delta Tech",    99999999
)

industry_toy <- tribble(
    ~gsubind,  ~subind,
    15104020,  "Diversified Metals & Mining",
    30101030,  "Food Retail",
    10102020,  "Oil & Gas Exploration & Production",
    20305020,  "Highways & Railtracks"
)

firms_toy
industry_toy
```

**(a)** Look carefully at the two tables before running any joins.

* Which firm has a `gsubind` that does **not** appear in `industry_toy`?
* Which subindustry has no firm in `firms_toy`?

Now **predict**, without running anything, how many rows each of these will return:

* `left_join(firms_toy, industry_toy, by = join_by(gsubind))`
* `inner_join(firms_toy, industry_toy, by = join_by(gsubind))`
* `anti_join(firms_toy, industry_toy, by = join_by(gsubind))`

Write your three predictions down before you continue.

**(b)** Now run all three joins and compare the results with your predictions.

```{r}
#| eval: false
firms_toy |> join_one(industry_toy, by = join_by(gsubind))

firms_toy |> join_two(industry_toy, by = join_by(gsubind))

firms_toy |> join_three(industry_toy, by = join_by(gsubind))
```

For each join, say what happened to Delta Tech and why.

**(c)** What happens if we swap the table on the left with the table on the right for each join?

```{r}
#| eval: false
industry_toy |> join_one(firms_toy, by = join_by(gsubind))

industry_toy |> join_two(firms_toy, by = join_by(gsubind))

industry_toy |> join_three(firms_toy, by = join_by(gsubind))
```

What does this result tell you? Which joins are directional?

**(d)** Complete the sentences below in your own words.

* I would use `left_join()` when I want to _____
* I would use `inner_join()` when I want to _____
* I would use `anti_join()` when I want to _____

**(e)** So far each firm has appeared only once. Real firm-year data is different, because the same firm appears in several years.

Run the code below to create a second pair of tiny tables.

```{r}
prices_toy <- tribble(
    ~gvkey, ~fyear, ~price,
    "001",  2022,   10.0,
    "001",  2023,   12.0,
    "001",  2024,   15.0
)

pe_toy <- tribble(
    ~gvkey, ~fyear, ~pe,
    "001",  2023,   18.0,
    "001",  2024,   20.0
)
```

Now join them using **only** `gvkey`.

```{r}
#| eval: false
prices_toy |> left_join(pe_toy, by = join_by(gvkey))
```

* How many rows did you expect? How many did you get?
* R printed a warning. What does it say?
* Look at the column names in the result. What happened to `fyear`?
* Look at the row for 2022. Which P/E ratios has it been matched to, and is that sensible?

**(f)** Now join the same two tables using **both** `gvkey` and `fyear`.

```{r}
#| eval: false
prices_toy |> left_join(pe_toy, by = join_by(gvkey, fyear))
```

* How many rows now?
* Which row has a missing `pe`, and why is that the right answer?
* In one sentence: why did the join in part (e) need a second key?

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Predict the output of a join before running it, then check that prediction.

* Distinguish `left_join()`, `inner_join()` and `anti_join()` by what each one keeps and drops.

* Recognise that `left_join()` and `anti_join()` ask directional questions, so swapping the tables asks something different.

* See a join key that is too coarse produce duplicated rows, at a scale where the duplication can be counted by hand.

✅ **Core Concepts to Highlight**

The three joins differ only in which rows survive. `left_join()` keeps everything on the left, `inner_join()` keeps only matches, `anti_join()` keeps only non-matches.

`left_join()` and `anti_join()` are directional and swapping the tables changes the question being asked. `inner_join()` is not directional.

A join key must identify the observational unit. In part (e) `gvkey` identifies a firm but not a firm-year, so each price row matches every P/E row for that firm.

The `fyear.x` / `fyear.y` columns in part (e) are the visible symptom: because `fyear` was not part of the key, R had no reason to line the two year columns up, so it kept both.

Row counts are a cheap diagnostic. Students should get into the habit of checking them after any join.

💬 **Suggested In-Class Prompts**

*"Before you run it — how many rows? Commit to a number."*

*"Delta Tech survived one join and vanished in another. Which, and why?"*

*"What business question does each `anti_join()` direction answer?"*

*"The 2022 price got a 2023 P/E ratio. What would that do to an analysis?"*

*"How would you have spotted this problem if the table had 9,000 rows instead of 3?"*

📌 **Common Misunderstandings**

Students may skip the prediction in part (a) and go straight to running the code. The prediction is the point; ask them to write numbers down before they type.

Students may expect `anti_join()` to return the matching rows, or to return columns from the right-hand table. It returns neither — only unmatched rows from the left, with the left table's columns.

Students may assume the two `anti_join()` directions are interchangeable.

Students may read the many-to-many warning as an error and stop, rather than as a diagnostic about their key.

Students may think 6 rows is simply "more data" rather than the same three observations duplicated.

⏱️ **Timing**

**25 minutes.** Prioritise understanding of parts (d) and (e). Compress (c) if running behind.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

Delta Tech (`gvkey` 004) has `gsubind` 99999999, which does not appear in `industry_toy`.

Highways & Railtracks (`gsubind` 20305020) has no firm in `firms_toy`.

The predictions should be:

* `left_join()`: 4 rows — every firm is kept
* `inner_join()`: 3 rows — Delta Tech is dropped
* `anti_join()`: 1 row — only Delta Tech

**(b).**

```{r}
firms_toy |> left_join(industry_toy, by = join_by(gsubind))
```

```{r}
firms_toy |> inner_join(industry_toy, by = join_by(gsubind))
```

```{r}
firms_toy |> anti_join(industry_toy, by = join_by(gsubind))
```

Delta Tech is treated differently by each join:

* `left_join()` keeps Delta Tech but leaves `subind` as `NA`, because there was nothing to fill it with.
* `inner_join()` drops Delta Tech entirely, because it has no match in `industry_toy`.
* `anti_join()` returns only Delta Tech, because it is the only firm without an industry match.

Note also that `anti_join()` returns just the columns from `firms_toy`. It is not adding information, it is filtering.

**(c).**

```{r}
industry_toy |> left_join(firms_toy, by = join_by(gsubind))
```
Compared with our original `left_join()`:

* We lose Delta Tech because it doesn't exist in the left table (`industry_toy`)
* We keep the Highways & Railtracks industry, but `gvkey` and `conm` are `NA` because there was nothing to fill it with from the `firms_toy` table.

```{r}
industry_toy |> inner_join(firms_toy, by = join_by(gsubind))
```

The same three firms are returned, so `inner_join()` is not directional — only rows that match in both tables survive, whichever table you start from. The only difference is the column order, which follows whichever table came first.

```{r}
industry_toy |> anti_join(firms_toy, by = join_by(gsubind))
```

This returns Highways & Railtracks — a subindustry with no firms in our list. In contrast, the original `anti_join()` returned Delta Tech — a firm with no subindustry in our list.

Overall, we see that `left_join()` and `anti_join()` are directional, while `inner_join()` is not. `left_join()` keeps everything from the left table, and `anti_join()` returns only rows from the left table that have no match in the right table. Swapping the tables changes which rows are kept or dropped.

**(d).**

Reasonable answers:

* `left_join()` — when I want to add information to a table without losing any of its rows.
* `inner_join()` — when I want to keep only the observations that appear in both tables.
* `anti_join()` — when I want to find the rows that failed to match, usually to check data quality.

**(e).**

```{r}
#| warning: true
prices_toy |> left_join(pe_toy, by = join_by(gvkey))
```

The result has **6 rows**, not 3. Each of the three price rows matched both of the P/E rows, because `gvkey` alone does not distinguish the years.

R warns about an "unexpected many-to-many relationship", telling us that rows on each side matched multiple rows on the other. This is a warning rather than an error, so the code still runs and produces a misleading dataset.

The clearest symptom is in the column names: the result contains `fyear.x` and `fyear.y` rather than a single `fyear`. Because `fyear` was not part of the join key, R had no reason to treat the two year columns as the same variable, so it kept both and disambiguated them.

Looking at 2022: its price of 10.0 has been paired with both the 2023 P/E ratio (18.0) and the 2024 P/E ratio (20.0). Neither is sensible. The 2022 row should have no P/E at all, because `pe_toy` does not cover 2022.

**(f).**

```{r}
prices_toy |> left_join(pe_toy, by = join_by(gvkey, fyear))
```

Now there are **3 rows**, matching the three rows we started with.

The 2022 row has a missing `pe`, which is the right answer: `pe_toy` genuinely has no 2022 observation, so there is nothing to attach. A missing value here is honest, whereas the 6-row result in part (e) invented matches that do not exist.

The join in part (e) needed a second key because one row in these tables represents one **firm-year**, not one firm, so it takes both `gvkey` and `fyear` together to identify an observation.

:::

:::

:::

<!-- END PROFILE:r-solutions -->
