<!-- question-type: inclass -->
### Exercise 4: Adding P/E Ratios with a Join

**(a)** We want to add P/E ratios to our tidy ASX firm-year dataset.

Run the code below to load and prepare the P/E ratio data.

```{r}
#| eval: false
pe_data <-
    read_csv("data/pe.csv") |>
    select(gvkey, fyear, pe) |>
    filter(!is.na(pe)) |>
    arrange(gvkey, fyear)
```

What variables in `pe_data` seem to identify each row?

**(b)** In `asx_tidy`, what does one row represent?

Which variable or variables together identify a unique observation in this dataset?

**(c)** Which **two** variables should we use to join `asx_tidy` and `pe_data`?

**(d)** Why do we need both of these variables in the join?

**(e)** Complete the code below to add the `pe` variable from `pe_data` to `asx_tidy`.

```{r}
#| eval: false
asx_with_pe <-
    asx_tidy |>
    left_join(
        pe_data,
        by = YOUR_CODE_HERE
    )
```

**(f)** After the join, what changed in the dataset?

In your answer, explain:

* what new information was added
* whether the join mainly added rows or columns

**(g)** Now suppose we join the two datasets using only `gvkey`.

Run the code below.

```{r}
#| eval: false
asx_wrong_join <-
    asx_tidy |>
    left_join(
        pe_data,
        by = YOUR_CODE_HERE
    )
```

What goes wrong here, and why?

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Recognise that after tidying, the observational unit in `asx_tidy` is a firm-year rather than just a firm.

* Identify when a join requires multiple keys rather than a single key.

* Use `left_join()` with `join_by()` to combine two firm-year datasets.

* Explain what can go wrong when a join is done with an incomplete key.

✅ **Core Concepts to Highlight**

Observational unit: after Exercise 3, each row in `asx_tidy` represents one firm in one year.

A correct join key must match the observational unit. Here, `gvkey` identifies the firm, while `fyear` identifies the year, so both are needed together.

A `left_join()` keeps the rows from the left-hand dataset and adds matching columns from the right-hand dataset.

Using too few join keys can create incorrect duplication because one observation can match many rows in the other table.

💬 **Suggested In-Class Prompts**

* “What does one row mean now that the data are tidy?”*

* “Why is `gvkey` not enough on its own once firms appear in multiple years?”*

* “What should stay the same after the correct join: rows, columns, or both?”*

* “What kind of error are we making when we join only on firm and ignore year?”*

📌 **Common Misunderstandings**

Students may say that `gvkey` alone identifies a row because it identifies the firm, forgetting that the same firm appears in multiple years.

Students may know both variables are needed but still struggle to explain why in terms of the observational unit.

Some students may think a join is successful whenever it runs, even if it duplicates rows incorrectly.

Students may not initially connect an incorrect join to downstream analytical errors.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

The variables that seem to identify each row are `gvkey` and `fyear`.

That is because each row appears to represent one firm in one year, with `pe` giving the P/E ratio for that firm-year observation.

**(b).**

One row in `asx_tidy` represents one firm in one year.

So the variables that together identify a unique observation are `gvkey` and `fyear`.

**(c).**

The two variables we should use are `gvkey` and `fyear`.

**(d).**

We need both variables because `gvkey` tells us which firm we are looking at, but it does not tell us which year. Since each firm appears in multiple years, joining only on `gvkey` would not uniquely identify the correct match. Adding `fyear` makes the match happen at the firm-year level, which is the correct observational unit in both datasets.

**(e).**

```{r}
asx_with_pe <-
    asx_prices_tidy |>
    left_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

**(f).**

The join added a new variable, `pe`, to the dataset.

It mainly added a column rather than rows, because we matched one firm-year observation in `asx_tidy` to one firm-year observation in `pe_data`.

**(g).**

```{r}
asx_wrong_join <-
    asx_prices_tidy |>
    left_join(
        pe_data,
        by = join_by(gvkey)
    )
```

This goes wrong because each firm in `asx_tidy` can now match multiple years in `pe_data` for the same firm. That creates duplicated rows and incorrect matches across years.

For example, a 2023 observation for one firm could be matched not only to its 2023 P/E ratio, but also to that firm’s 2022 or 2024 P/E ratios. So the join runs, but the resulting dataset is no longer correctly aligned at the firm-year level.

:::

:::

:::

<!-- END PROFILE:r-solutions -->

