<!-- question-type: prepare -->
### Exercise 1: Identifying Issues with Untidy Data

We will start with a deliberately messy version of the ASX stock price data. Run the following code to load the dataset.

```{r}
#| eval: false
asx_prices_messy <- read_csv("data/asx_prices_messy.csv")
```

Then inspect a subset of the data:

```{r}
asx_prices_messy |>
    select(gvkey, conm, price_2023, price_2024,
           eps_2023, eps_2024) |>
    head(10)
```

**(a)** In what ways is this data frame not tidy? What principles of tidy data are violated?

**(b)** Use the starter code below to compute the average share price across all firms for 2023 and 2024.

```{r}
#| eval: false

asx_prices_messy |>
    summarise(
        # Add up all 2023 share prices
        sum_prices_2023 = sum(YOUR_CODE_HERE, na.rm = TRUE),
        # Add up all 2024 share prices
        sum_prices_2024 = sum(YOUR_CODE_HERE, na.rm = TRUE),
        # Count how many firms have a non-missing share price in 2023
        count_firms_2023 = sum(!is.na(price_2023)),
        # Count how many firms have a non-missing share price in 2024
        count_firms_2024 = sum(!is.na(price_2024))
    ) |>
    mutate(
        # Combine the 2023 and 2024 price totals
        total_sum = YOUR_CODE_HERE + YOUR_CODE_HERE,
        # Combine the number of non-missing firm prices across both years
        total_count = count_firms_2023 + count_firms_2024
    ) |>
    mutate(
        # Divide the total sum of prices by the total number of firms
        avg_price_23_24 = YOUR_CODE_HERE / YOUR_CODE_HERE
    ) |> 
    select(avg_price_23_24)
```

**(c)** Why does the approach in part (b) not scale well as more years are added?

**(d)** The code below plots share price against earnings per share in 2024.

```{r}
#| warning: false
#| eval: false

asx_prices_messy |>
    filter(eps_2024 > -5) |>
    ggplot(aes(x = eps_2024, y = price_2024)) +
    YOUR_CODE_HERE +
    YOUR_CODE_HERE +
    labs(
        title = "YOUR_LABEL_HERE",
        x = "YOUR_LABEL_HERE",
        y = "YOUR_LABEL_HERE"
    ) +
    theme_minimal()
```

What relationship do you see? Why does this make sense economically?

**(e)** Suppose you also wanted to include 2023 in this analysis. Briefly explain:

* which extra columns you would need
* what you would need to change in the plotting process
* why this approach would become awkward if extended to 2019 to 2024

**(f)** In two or three sentences, explain to a manager why this data structure makes analysis harder than it needs to be.

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Identify and explain why the provided data is not tidy.
* Practise working with a wide data structure and see why it becomes awkward for repeated analysis tasks.
* Build intuition for why analysts often reshape data before doing summaries or visualisations.
* Communicate, in plain language, why data structure matters for business analysis.

✅   **Core Concepts to Highlight**

* Tidy data rules: one variable per column, one observation per row.
* Wide data often store the same variable across many year-specific columns.
* A structure that works for viewing data can still be awkward for analysis.
* Scalability: code that works for one or two years may become repetitive and brittle as more years are added.
* Analysts often tidy data to make summaries, plots, and comparisons easier to repeat.

💬 **Suggested In-Class Prompts** (if needed)

* What exactly is stored in the column names here?
* Why is part (b) more cumbersome than it should be?
* What would happen if we added another five years of data in this format?
* Why might a manager care about the way the data are structured if the numbers themselves have not changed?

📌 **Common Misunderstandings**

* Thinking that “messy” just means visually untidy rather than structurally awkward.
* Saying the data are not tidy without explaining which tidy-data principles are violated.
* Treating the code in part (b) as a reasonable long-run solution rather than a short-term workaround.
* Explaining part (e) vaguely, without linking the difficulty to repeated year-specific columns.
* Giving a managerial explanation that focuses on R syntax rather than on speed, flexibility, and ease of analysis.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a)** This data frame is not tidy because the year is stored in the column names rather than in its own variable. Columns such as `price_2023` and `price_2024` contain values of the variable `year`, which should instead appear in a separate column. The data also mix two variables, the measure (`price` or `eps`) and the year, inside a single column name. More broadly, one firm-year observation is spread across multiple columns rather than appearing as one row.

**(b)**

```{r}
asx_prices_messy |>
    summarise(
        # Add up all 2023 share prices
        sum_prices_2023 = sum(price_2023, na.rm = TRUE),
        # Add up all 2024 share prices
        sum_prices_2024 = sum(price_2024, na.rm = TRUE),
        # Count how many firms have a non-missing share price in 2023
        count_firms_2023 = sum(!is.na(price_2023)),
        # Count how many firms have a non-missing share price in 2024
        count_firms_2024 = sum(!is.na(price_2024))
    ) |>
    mutate(
        # Combine the 2023 and 2024 price totals
        total_sum = sum_prices_2023 + sum_prices_2024,
        # Combine the number of non-missing firm prices across both years
        total_count = count_firms_2023 + count_firms_2024
    ) |>
    mutate(
        # Divide the total sum of prices by the total number of firms
        avg_price_23_24 = total_sum / total_count
    ) |> 
    select(avg_price_23_24)
```

**(c)** This approach does not scale well because every new year adds more columns that must be handled manually. We would need to keep adding extra sums, counts, and references to year-specific variables, making the code longer and easier to break.

**(d)**

```{r}
asx_prices_messy |>
    filter(eps_2024 > -5) |>
    ggplot(aes(x = eps_2024, y = price_2024)) +
    geom_point(alpha = 0.4) +
    geom_smooth(method = "lm", se = TRUE) +
    labs(
        title = "Share Price vs EPS (2024)",
        x = "Earnings per Share (EPS, 2024)",
        y = "Share Price (2024)"
    ) +
    theme_minimal()
```

The plot suggests a positive relationship between earnings per share and share price in 2024. This makes sense economically because investors are buying a claim on the firm’s future earnings, so firms with stronger earnings tend to attract more demand and higher share prices.

**(e)** A good answer would explain that we would need to bring in the 2023 columns, such as `eps_2023` and `price_2023`, in addition to the 2024 ones. The plotting process would become more complicated because we would need to manually combine or stack information from multiple years before plotting. This would become increasingly awkward from 2019 to 2024 because each extra year would add more column references and more manual work, making the analysis repetitive and harder to maintain.

**(f)** A manager-friendly explanation might be:

> The current structure makes even simple comparisons across years more cumbersome than they need to be. Reorganising the data would make it easier to summarise trends, build plots, and update the analysis when new years of data are added.

:::

:::

:::

<!-- END PROFILE:r-solutions -->
