### Exercise 6: Auditing Missing P/E Ratios Before an Analytics Meeting

Suppose your team is preparing for an analytics meeting. You have built a firm-year dataset with share prices, earnings, and P/E ratios, but many firm-year observations still have missing `pe` values.

Before moving ahead, you want to understand:

* what the full dataset looks like
* what the complete-match dataset looks like
* which observations are missing P/E
* what you should ask the data team before continuing

**(a)** In Exercise 4, we used a `left_join()` to add P/E ratios from `pe_data` to `asx_tidy`.

What happened to the rows in `asx_tidy` that did **not** have a matching P/E ratio in `pe_data`? Briefly explain what the join did in those cases.

**(b)** Run the code below, which uses a different kind of join, called `inner_join()`.

```{r}
#| eval: false
asx_with_pe_inner <-
    asx_prices_tidy |>
    inner_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

How is `asx_with_pe_inner` different from `asx_with_pe`? What does `inner_join()` keep that `left_join()` does not?

**(c)** Run the code below, which uses another kind of join, called `anti_join()`.

```{r}
#| eval: false
missing_pe <-
    asx_prices_tidy |>
    anti_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

What does `missing_pe` represent? How is this different from the dataset created with `inner_join()` in part (b)?

**(d)** Compare the number of rows in the following datasets.

```{r}
#| eval: false
nrow(asx_prices_tidy)
nrow(asx_with_pe)
nrow(asx_with_pe_inner)
nrow(missing_pe)
```

What do you notice? What does this tell you about how much of the firm-year dataset has valid P/E coverage?


**(e)** Look at the observations in `missing_pe`.

Use code to investigate whether missing P/E ratios seem to be concentrated in certain years or associated with certain EPS values.
Write one or two short findings based on what you see.

**(f)** The `pe_data` file was provided to us as a separate dataset.

Why might some firm-year observations in `asx_tidy` still fail to find a matching P/E ratio in that file, even when `price` and `eps` are present?

In your answer, think about possible differences in:

* which firms were included
* which years were covered
* how P/E ratios were defined or filtered before the file was given to us

**(g)** Suppose you are preparing for a meeting with the data team before using `pe` in your analysis.

What are two or three specific questions you would ask to check whether the missing P/E values reflect:

* a normal feature of the data
* a difference in firm or year coverage
* a problem in how the file was sourced or matched

If the missing values seem to come from incomplete firm coverage, what might your team do next to source the missing data?

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Explain what happens to unmatched rows under a `left_join()`.

* Interpret the outputs of `inner_join()` and `anti_join()` when auditing data coverage.

* Compare full-sample, matched-sample, and unmatched observations in a firm-year dataset.

* Use join outputs to think critically about missing data before proceeding with analysis.

✅ **Core Concepts to Highlight**

A `left_join()` keeps all rows from the left-hand dataset and adds matching values where available. Unmatched rows stay in the dataset with missing values.

An `inner_join()` keeps only rows that have a match in both datasets.

An `anti_join()` keeps only rows from the left-hand dataset that do not have a match in the right-hand dataset.

These joins are not only tools for combining data. They are also tools for checking data readiness and diagnosing missing coverage.

💬 **Suggested In-Class Prompts**

*“What is the difference between a missing value after a left join and a row disappearing in an inner join?”*

*“If you only analysed the inner-join sample, what observations would you be throwing away?”*

*“What kinds of business or data-construction reasons could explain missing P/E ratios?”*

*“What would you want to clarify with the data team before using P/E in a report?”*

📌 **Common Misunderstandings**

Students may think `inner_join()` is simply a ‘better’ join because it produces complete data, without noticing that it drops observations.

Students may confuse missing `pe` values with a coding error, when in practice they may reflect real data limitations or variable-definition issues.

Students may not initially see that `missing_pe` comes from the left-hand dataset, not from `pe_data`.

Students may overlook that dropping unmatched observations could bias later analysis if the missingness is systematic.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a).**

Rows in `asx_tidy` that did not have a matching P/E ratio in `pe_data` were still kept in the dataset.

The `left_join()` added `pe` where a match was found, and left `pe` as missing (`NA`) where no match was found.

**(b).**

```{r}
asx_with_pe_inner <-
    asx_prices_tidy |>
    inner_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

`asx_with_pe_inner` contains only the firm-year observations that have a match in both `asx_tidy` and `pe_data`.

Compared with `asx_with_pe`, it drops the rows where no matching P/E ratio was found.

So `inner_join()` keeps only complete matches, while `left_join()` keeps all rows from `asx_tidy` whether they match or not.

**(c).**

```{r}
missing_pe <-
    asx_prices_tidy |>
    anti_join(
        pe_data,
        by = join_by(gvkey, fyear)
    )
```

`missing_pe` contains the firm-year observations in `asx_tidy` that did not find a match in `pe_data`.

This is different from the `inner_join()` result because `inner_join()` keeps the matched observations, while `anti_join()` keeps the unmatched ones.

**(d).**

`asx_tidy` and `asx_with_pe` should have the same number of rows, because the `left_join()` keeps every row from `asx_tidy`.

`asx_with_pe_inner` should have fewer rows if some firm-year observations do not have a matching P/E ratio.

`missing_pe` shows exactly the observations that were left out of the inner join.

Together, these row counts show how much of the firm-year dataset has valid P/E coverage and how much is missing.

**(e).**

Students’ exact answers will vary, but a good answer would identify one or two patterns in `missing_pe`.

For example:

* missing P/E ratios may be concentrated in certain years
* missing P/E ratios may occur more often when EPS is negative or very close to zero
* missing P/E rations for certain firms

Illustrative code could include:

```{r}
missing_pe |>
    count(fyear)
```

```{r}
missing_pe |>
    summarise(
        min_eps = min(eps, na.rm = TRUE),
        max_eps = max(eps, na.rm = TRUE),
        avg_eps = mean(eps, na.rm = TRUE)
    )
```

```{r}
missing_pe |>
  count(conm) |>
  arrange(desc(n))
```


**(f).**

Possible explanations include:

* The supplied `pe_data` file may not cover every firm in our ASX dataset.
* Some firms may appear in `asx_tidy` but be missing from `pe_data` for certain years.
* The P/E file may cover a different time period from the main ASX dataset.
* P/E ratios may be intentionally omitted when EPS is negative or very close to zero.
* The data team may have filtered the file before giving it to us, so only a subset of firm-years was included.
* The two datasets may have been sourced under different rules, even if they use the same identifiers.

**(g).**

Possible questions for the data team include:

* Does `pe_data` cover all firms in our ASX dataset, or only a subset?
* Are some years intentionally missing from the P/E file?
* How is P/E defined in this dataset, and are firms with negative or near-zero EPS excluded?
* Was the P/E file filtered before it was given to us?
* Should missing P/E values be treated as expected missingness, or as a sign that the match or source file needs to be checked?

If the missing values seem to reflect incomplete firm coverage, the team might:

* check whether the missing firms can be sourced from the original provider
* request a broader extract from the data team
* look for an additional source covering those firms and years
* decide whether the analysis should be restricted to firms with reliable P/E coverage

:::

:::

:::

<!-- END PROFILE:r-solutions -->
