<!-- question-type: prepare -->
### Exercise 2: Combining Datasets

In this exercise, you will combine information from two datasets using a join. Start with a quick look at the `firm_codes` and `subindustry_names` datasets:

```{r}
firm_codes |>
    head(10)
```

```{r}
subindustry_names |>
    head(10)
```

**(a)** Look at the variables in both datasets. Which variable or variables look like they could be used to join these two datasets?

**(b)** Which join variable would you choose? Explain why it is the best choice here.

**(c)** Use a left join to add the subindustry name to each firm in `firm_codes`.

```{r}
#| eval: false

firm_info <-
    YOUR_DATASET_NAME |>
    left_join(YOUR_DATASET_NAME, by = join_by(YOUR_VARIABLE_NAME))

firm_info |>
    head(10)
```

**(d)** What changed after the join? Briefly explain:

* what information was added
* whether the join mainly added rows or columns
* why this makes the dataset more useful for analysis

**(e)** Look at how `gsubind` behaves in the two datasets.

* In which dataset would you expect each `gsubind` to appear only once?
* In which dataset can the same `gsubind` appear many times?
* Why does that make sense?

**(f)** Fill in the variable name in each sentence below.

In `subindustry_names`, `_____` behaves like a **primary key** because each code identifies one subindustry label.

In `firm_codes`, `_____` behaves like a **foreign key** because many firms can share the same subindustry code.

**(g)** Suppose we reversed the join and started with `subindustry_names` on the left-hand side. What would be different about the resulting dataset, and why?

<!-- BEGIN PROFILE:r-teaching-guide -->

::: {.content-visible when-profile="r-teaching-guide"}

::: {.teaching-block}

::: {.teaching-block-header}
Teaching Note
:::

::: {.teaching-block-body}

🎯 **Learning Objective**
Students should:

* Identify a sensible join key by comparing the variables in two tables.
* Use a `left_join()` to combine a firm-level dataset with a lookup table.
* Explain how the role of a variable can differ across tables.
* Build intuition for why the table on the left-hand side matters in a join.

✅   **Core Concepts to Highlight**

* A join uses a shared variable, called a key, to match rows across tables.
* Not every shared-looking variable is a good key. Students should think about reliability and level of match.
* A `left_join()` keeps all rows from the table on the left and adds matching information from the table on the right.
* In a lookup table, a variable may uniquely identify rows. In another table, the same variable may be used repeatedly to link observations back to those rows.
* This is a natural place to introduce the language of **primary key** and **foreign key**.

💬 **Suggested In-Class Prompts** (if needed)

* Which variables appear in both tables, and which one would you trust more for matching?
* Why is `gsubind` a better choice than something like a name column?
* After the join, what stayed the same and what changed?
* Why does the left-hand table matter in a `left_join()`?
* What role is `gsubind` playing in each table?

📌 **Common Misunderstandings**

* Assuming that any variable that appears in both tables is equally good for joining.
* Thinking that a join always changes the number of rows.
* Describing the result of a `left_join()` without noting that the left-hand table determines which rows are preserved.
* Treating primary key and foreign key as abstract jargon rather than as descriptions of a variable’s role in a table.

:::

:::

:::

<!-- END PROFILE:r-teaching-guide -->

<!-- BEGIN PROFILE:r-solutions -->

::: {.content-visible when-profile="r-solutions" when-profile="r-teaching-guide"}

::: {.solution-block}

::: {.solution-block-header}
Solution
:::

::: {.solution-block-body}

**(a)** The most plausible join variable is `gsubind`. Depending on the columns shown in `subindustry_names`, students may also mention other code-like variables, but `gsubind` is the key variable we want here because it appears in both tables and represents the subindustry code.

**(b)** `gsubind` is the best choice because it is a structured code variable designed to represent the same subindustry in both tables. In general, numeric or code-based match keys are usually safer than text variables because they are more standardised and less likely to vary across datasets. A text variable like `conm` would be a worse choice because company names can be abbreviated, misspelled, formatted differently, or written inconsistently across files. So the key point is not just that `gsubind` fits the business meaning here, but that coded identifiers are often much more reliable join keys than text labels.

**(c)**

```{r}
firm_info <-
    firm_codes |>
    left_join(subindustry_names, by = join_by(gsubind))

firm_info |>
    head(10)
```

**(d)** The join adds subindustry names to the firm-level dataset. It mainly adds columns rather than rows, because we are attaching descriptive information to each existing firm. This makes the dataset more useful because we can now analyse firms using meaningful industry labels rather than just codes.

**(e)** In `subindustry_names`, we would expect each `gsubind` to appear only once because each code should map to one subindustry label. In `firm_codes`, the same `gsubind` can appear many times because many firms can belong to the same subindustry. That makes sense because `subindustry_names` is a lookup table, while `firm_codes` is a firm-level table.

**(f)** In both sentences, the missing variable name is `gsubind`.

In `subindustry_names`, `gsubind` behaves like a **primary key** because each code identifies one subindustry label.

In `firm_codes`, `gsubind` behaves like a **foreign key** because many firms can share the same subindustry code.

**(g)** If we started with `subindustry_names` on the left-hand side, the resulting dataset would keep all subindustry rows rather than all firm rows. That means the structure of the output would be organised around subindustries, not firms. In a `left_join()`, the table on the left determines which rows are preserved.

```{r}
subindustry_first <-
    subindustry_names |>
    left_join(firm_codes, by = join_by(gsubind))

subindustry_first |>
    head(10)
```

This version starts from the subindustry lookup table and then brings in the firms that match each subindustry code. The result is therefore anchored on subindustries rather than firms, which is different from the earlier join where every firm row was preserved first.

:::

:::

:::

<!-- END PROFILE:r-solutions -->
