## The Business Challenge

```{r}
#| warning: false
#| echo: false
library(tidyverse)
```

```{r}
#| warning: false
#| eval: true
#| echo: false
#| cache: true

# Create data/ directory if it doesn't exist
if (!dir.exists("data")) {
  dir.create("data")
}

# Define URL and destination file path
url <- "https://raw.githubusercontent.com/unimelb-cmce-10002/fba-book/refs/heads/main/data/asx_prices_messy.csv"
destfile <- "data/asx_prices_messy.csv"
# Download the file if it doesn't already exist
if (!file.exists(destfile)) {
  download.file(url, destfile, mode = "wb")
}
# Load the data
asx_prices_messy <- read_csv(destfile)

# Define URL and destination file path
url <- "https://raw.githubusercontent.com/unimelb-cmce-10002/fba-book/refs/heads/main/data/asx_200_2024.csv"
destfile <- "data/asx_200_2024.csv"
# Download the file if it doesn't already exist
if (!file.exists(destfile)) {
  download.file(url, destfile, mode = "wb")
}
# Load the data
asx_200_2024 <- read_csv(destfile)
# Wrangle
firm_codes <-
    asx_200_2024 |>
    select(gvkey, conm, gsubind)

# Define URL and destination file path
url <- "https://raw.githubusercontent.com/unimelb-cmce-10002/fba-book/refs/heads/main/data/GICS_subindustry.csv"
destfile <- "data/GICS_subindustry.csv"
# Download the file if it doesn't already exist
if (!file.exists(destfile)) {
  download.file(url, destfile, mode = "wb")
}
# Load the data
subindustry_names <- read_csv(destfile)

# Define URL and destination file path
url <- "https://raw.githubusercontent.com/unimelb-cmce-10002/fba-book/refs/heads/main/data/pe.csv"
destfile <- "data/pe.csv"
# Download the file if it doesn't already exist
if (!file.exists(destfile)) {
  download.file(url, destfile, mode = "wb")
}
# Load the data
pe_data <- read_csv(destfile)
# Wrangle data
pe_data <-
    pe_data |>
    select(gvkey, fyear, pe) |>
    filter(!is.na(pe)) |>
    arrange(gvkey, fyear)
```

### The Topic: What shapes how the market values Australian firms? {.unnumbered}

Investors do not just care about whether a company is profitable. They also care about how much they are paying for those earnings.

In this tutorial, we will work with data on large Australian firms to explore a simple question:

**How strongly are share prices and valuation ratios related to company earnings, and how do those patterns vary across firms and industries?**

To answer that question, we first need to prepare the data. Some of the financial variables are stored in a messy wide format, and other information is spread across separate files. So before we can analyse anything, we need to:

* reshape the data into tidy form
* join together firm, industry, and valuation data
* check what information is missing after those joins

Two key variables in today’s tutorial are:

$$
\text{EPS} = \frac{\text{Profit}}{\text{Shares Outstanding}}
$$

EPS tells us how much profit a company earns for each share.

$$
\text{P/E} = \frac{\text{Share Price}}{\text{EPS}}
$$

The P/E ratio tells us how much investors are willing to pay for each dollar of earnings.

By the end of the tutorial, you will have built and explored a dataset that lets you compare firms and industries using share prices, earnings, and valuation ratios.

## R packages for today {.unnumbered}

```{r}
#| eval: false
library(tidyverse)     # collection of packages for data manipulation and visualization
```
